---
title: "Original Analysis on Hub Detection"
author: "Jose Angel Sanchez Gomez"
date: '2023-03-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

df_reduced <- read.csv("312_df_reduced.csv")
head(df_reduced[, 1:5])
```




## Original Analysis

To observe what are the results when controlling for underlying clinical variables, lets first re-visualize the results without controlling for these variables.

```{r, fig.width = 10, fig.height = 5}

df_reduced <- df_reduced[, -1]
df_genes <- df_reduced[, 7 + 1:200]
cor <- cor(df_genes)

eig <- eigen(cor)

par(mfrow = c(1,3))
plot(1 / eig$values, main = "Precision Matrix: Eigenvalues")
plot(eig$values[200:100 - 1] / eig$values[200:100], main = "Precision Matrix: Eigenvalue Ratios")
abline(v = 3, col = "red")

inf_measure <- apply(eig$vectors[, 198:200]^2, MARGIN = 1, sum)
plot(inf_measure, main = "Influence Measures")
abline(h = 2 * 3 / 200)
colnames(df_genes[, which(inf_measure > 2 * 3 / 200)])


rm("df_genes", "cor", "eig", "inf_measure")

```

As we can observe in our original results, without controlling for underlying clinical variables, we obtain 3 separating eigenvalues, and 5 hub variables: "SEMG1",   "RN7SK",   "SEMG2",   "MIR3609" and "SCARNA7". 



## Preliminary steps

Before jumping into our analysis, we will first ensure that our variables of interest are clean and ready to use.


```{r}

colnames(df_reduced[, 1:7])

df_reduced$race <- as.factor(df_reduced$race)
df_reduced$vital_status <- as.factor(df_reduced$vital_status)
df_reduced$primary_diagnosis <- as.factor(df_reduced$primary_diagnosis)
df_reduced$radiation_therapy <- as.factor(df_reduced$radiation_therapy)
df_reduced$pharmaceutical_therapy <- as.factor(df_reduced$pharmaceutical_therapy)

```


With this processing, lets save the results, and in the next Rmd, we do the analysis of hub detection when regressing our data against these 6 clinical variables of interest.

```{r}

print(ls())
write.csv(df_reduced, file = "322_df_reduced.csv")
rm(list = ls())

```